Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-12-09 manuscript/usses_water_ms.Rmd
more revisions; adds sensitivity analysis

To that end, here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (approximately $\pm50\%$) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between interannual variation in available soil moisture and interannual variation in ANPP, and if community dynamics underlie the ecosystem responses.
In particular, we are interested in the consistency of the soil moisture-ANPP relationship among treatments.
Is the relationship steeper under the drought treatment at low soil moisture?
Does the relationship saturate under the irrigation treatment at high soil moisture?
To answer these questions we fit a generalized linear mixed effects model to test whether the regressions differed among treatments.
We also analyzed community composition over time, allowing us to place our experimental results within the framework our scenarios (Fig. 1).

@@

## Data Analysis
Our main goal was to test whether the relationship between ANPP and soil moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site [@Blaisdell1958; @Dalgleish2011; @Adler2012], we chose to use cumulative volumetric water content from March through June as our metric of soil moisture (hereafter referred to as `VWC').
To achieve this goal, we fit a generalized linear mixed effects regression model with log(ANPP) as the response variable and VWC and treatment as fixed effects.
Plot and year of treatment were included as random effects to account for non-independence of observations, as described below.
We log-transformed ANPP to account for heteroscedasticity.
Both log(ANPP) and VWC were standardized to have mean 0 and unit variance before fitting the model [i.e., $(x_i - \bar{x})/\sigma_x$].

@@

@@

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP (Table 1; Fig. 3B).
The effect of soil moisture for each treatment is associated with high uncertainty, however, with 95\% Bayesian credible intervals that overlap zero (Table 1).
Although the parameter estimates for the effect of soil moisture overlap zero, the posterior distributions of the slopes all shrank and shifted to more positive values relative to the prior distributions (Fig. A2-2), which indicates the data did influence parameter estimates beyond the information from the uninformative priors.
Ecosystem functional response was similar among treatments (Table 1; Fig. 3B), but there is evidence that the slope for the irrigation treatment is less than the slope for the control treatment.
This evidence comes from interpreting the posterior distribution of the slope offset for the irrigation treatment, from which we calculate a \new{96}% one-tailed probability that the estimate is less than zero (Fig. 3A, right panel).

Community composition was similar among treatments.
The multidimensional space of community composition overlapped among treatments in all years and was equally dispersed in all years (Table 2; Fig. 4).
Community composition was also remarkably stable over time, with no evidence of divergence among treatments (Table 2; Fig. 4).

@@

The results of our five year experiment in a sagebrush steppe conform to two of our four predictions, depending on treatment.
Ecosystem functional response and community composition did not change under chronic drought (Fig. 3A, Fig. 4), representing the top left scenario in Fig. 1.
Ecosystem functional response under chronic irrigation was different from the control treatment, but community composition remained unchanged (Fig. 3A, Fig. 4), representing the top right scenario in Fig. 1.
The decrease in the slope of the VWC $\times$ productivity relationship in the irrigated plots is consistent with a weaker response of productivity to additional water in this ecosystem.
This suggests that if average soil moisture were pushed consistently higher than currently observed ambient conditions, there would be a weaker relationship between precipitation and productivity in this system.
Nonetheless, the slopes of the VWC $\times$ productivity relationship were similar among treatments (Table 1), at least at the level of irrigation we imposed.
We therefore conclude that ecosystem functional response is consistent and weak across all precipitation treatments.

The similarity of ecosystem functional response (Fig. 3) and community composition (Fig. 4) among treatments is surprising because grasslands generally, and sagebrush steppe specifically, are considered water-limited systems.
\new{For example,} @Huxman2004 and @Knapp2015 \new{showed that semi-arid sites are more sensitive to drought than mesic sites, and} @Wilcox2017 \new{found that semi-arid sites are particularly sensitive to irrigation treatments.}
\new{Based on these findings}, we expected ecosystem functional response, community composition, or both to change under our treatments.
Why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons; two are limitations of our study, and one is the life history traits of the species in our focal communities.
We first discuss the potential limitations of our study, and then discuss the biological explanation.

@@

@@

Our results suggest the species in our focal plant community are insensitive to to changes in precipitation regime. 
Such insensitivity can buffer species against precipitation variability in this semi-arid ecosystem, making them successful in the long run.
Longer, chronic precipitation alteration might reveal plant community shifts that we did not observe [e.g., @Wilcox2016]. 
For example, a long-term increase in water availability could allow species that do not bet hedge to gain prominence and dominate the ecosystem functional response.
\new{Time \emph{per se} may be especially relevant in our focal ecosystem where the perennial species are long-lived, meaning compositional turnover may take many more years than we report on here.}
Our results suggest compositional shifts would have the largest impact at high rainfall because the current community maintained consistent ecosystem functional response at very low water availability.

>>>>

To that end, here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (approximately $\pm50\%$) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between interannual variation in available soil moisture and interannual variation in ANPP, and if community dynamics underlie the ecosystem responses.
In particular, we are interested in the consistency of the soil moisture-ANPP relationship among treatments.
Is the relationship steeper under the drought treatment at low soil moisture?
Does the relationship saturate under the irrigation treatment at high soil moisture?
To answer these questions we fit a generalized linear mixed effects model to test whether the regressions differed among treatments.
We also analyzed community composition and \new{the sensitivity of ANPP to drought and irrigation treatments} over time, allowing us to place our experimental results within the framework our scenarios (Fig. 1).

@@

## Data Analysis

### Ecosystem functional response

Our main goal was to test whether the relationship between ANPP and soil moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site [@Blaisdell1958; @Dalgleish2011; @Adler2012], we chose to use cumulative volumetric water content from March through June as our metric of soil moisture (hereafter referred to as `VWC').
To achieve this goal, we fit a generalized linear mixed effects regression model with log(ANPP) as the response variable and VWC and treatment as fixed effects.
Plot and year of treatment were included as random effects to account for non-independence of observations, as described below.
We log-transformed ANPP to account for heteroscedasticity.
Both log(ANPP) and VWC were standardized to have mean 0 and unit variance before fitting the model [i.e., $(x_i - \bar{x})/\sigma_x$].

@@

### Sensitivity of ANPP over time
\new{Unfortunately, our data does not allow us to include the effect of time on ecosystem functional response in the model described above. 
This is because we lack sufficient within-year and within-treatment variation of soil moisture (i.e., within a year and treatment each plot shares the same value of VWC). 
Therefore, we conducted an independent analysis to determine if the sensitivity of ANPP to treatment changed over time.}

\new{We define `sensitivity' as $\frac{ \text{ANPP}_{\text{control}} - \text{ANPP}_{\text{treatment}} } { \text{VWC}_{\text{control}} - \text{VWC}_{\text{treatment}} }$, following} @Wilcox2017.
\new{This metric of sensitivity measures the response of ANPP standardized to the magnitude of the treatment. Because our treatment and control plots are not paired, we conducted this analysis by comparing each treatment plot to the mean of the control plots in each year. Then, independently for each treatment, we regressed sensitivity against year of treatment using the \texttt{lm()} function in R} [@R2016].
\new{This analysis also allows us to link particularly sensitive treatment years to changes in community composition.}

### Community composition over time

@@

\new{We conducted the above analysis for all species, and then conducted separate analyses for perennial species only and annual species only. Annual species have shorter life spans, so conducting separate analyses allows us to discover if community responses are restricted to short-lived species that might respond more quickly to altered precipitation regimes. Given the dominance of perennial species in our system, shifts in the annual plant community could be masked in the analysis of the full community.}

### Reproducibility

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP (Table 1; Fig. 3B).
The effect of soil moisture for each treatment is associated with high uncertainty, however, with 95\% Bayesian credible intervals that \new{broadly} overlap zero (Table 1).
Although the parameter estimates for the effect of soil moisture overlap zero, the posterior distributions of the slopes all shrank and shifted to more positive values relative to the prior distributions (Fig. A2-2), which indicates the data did influence parameter estimates beyond the information from the uninformative priors.
Ecosystem functional response was similar among treatments (Table 1; Fig. 3B), but there is evidence that the \new{slopes for the drought and} irrigation treatments are less than the slope for the control treatment.
This evidence comes from interpreting the posterior distributions of the slope offsets for the treatments.
\new{From these distributions,} we calculate a \new{91}% one-tailed probability that the estimate is less than zero for the irrigation treatment \new{and a 96\% one-tailed probability that the estimate is less than zero for the drought treatment} (Fig. 3A, right panel).

\new{Sensitivity of ANPP to irrigation was constant over the course of our experiment (Fig. x). Sensivitity of ANPP to drought, however, grew over time (\emph{P} = 0.0002; Fig. x)}.

Community composition was similar among treatments.
The multidimensional space of community composition overlapped among treatments in all years and was equally dispersed in all years (Table 2; Fig. 4).
Community composition was also remarkably stable over time, with no evidence of divergence among treatments (Table 2; Fig. 4).
\new{Splitting the analysis apart for perennials and annuals led to similar results.
The perennial community was stable over time and consistently similar among treatments (Fig. Ax-x).
There is some evidence that the annual community changed in response to treatment in two years (Fig. Ax-x), but these responses appear to come from very rare species since our analysis weights them equally (Appendix X).}

@@

The results of our five year experiment in a sagebrush steppe \new{conforms to one} of our four predictions.
Ecosystem functional response under chronic \new{drought and irrigation} was different from the control treatment, but community composition remained unchanged (Fig. 3A, Fig. 4), representing the top right scenario in Fig. 1.
The decrease in the slope of the VWC $\times$ productivity relationship in the irrigated plots is consistent with a weaker response of productivity to additional water in this ecosystem.
This suggests that if average soil moisture were pushed consistently higher than currently observed ambient conditions, there would be a weaker relationship between precipitation and productivity in this system.

\new{The decrease in the slope of the VWC $\times$ productivity relationship in the droughted plots is puzzling because it is counter to our expectation that the slope would be higher than the control plots.
However, we believe the significance of the slope offsets has more to do with noise than signal.
This is because the full distributions of} the slopes of the VWC $\times$ productivity relationship, \new{not just the offsets}, were similar among treatments (Table 1).
We therefore conclude that ecosystem functional response is consistent (\new{similar values}) and weak (\new{all broadly overlapping zero}) across all precipitation treatments.

The similarity of ecosystem functional response (Fig. 3) and community composition (Fig. 4) among treatments is surprising because grasslands generally, and sagebrush steppe specifically, are considered water-limited systems.
\new{For example,} @Huxman2004 and @Knapp2015 \new{showed that semi-arid sites are more sensitive to drought than mesic sites, and} @Wilcox2017 \new{found that semi-arid sites are particularly sensitive to irrigation treatments.}
\new{Based on these findings}, we expected ecosystem functional response, community composition, or both to change under our treatments.
Why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons. 
Two are limitations of our study, and one is the life history traits of the species in our focal communities.
We first discuss the potential limitations of our study, and then discuss the biological explanation.

@@

\new{We did find that ANPP became more sensitive to drought over time (Fig. x).
Thus, the difference between ANPP, standardize by the difference in soil moisture, in drought and control plots grew over time.
The latter years of our experiment were relatively wet, meaning that the drought treatment could have the largest absolute reduction of soil moisture in those years.
But our metric of sensitivity accounts for differences among years because ANPP differences is standardized by the difference in soil moisture within the year.
Therefore, it is likely that the increased sensitivity to drought over time is due to the cimulative impact of drought on the dominant species.
Ecosystem functional response may soon shift if sensitivity continues to increase.}

@@

Our results suggest the species in our focal plant community are insensitive to to changes in precipitation regime, \new{at least over the five years of our experiment}. 
Such insensitivity can buffer species against precipitation variability in this semi-arid ecosystem, making them successful in the long run.
Longer, chronic precipitation alteration might reveal plant community shifts that we did not observe [e.g., @Wilcox2016]. 
For example, a long-term increase in water availability could allow species that do not bet hedge to gain prominence and dominate the ecosystem functional response.
\new{Time \emph{per se} may be especially relevant in our focal ecosystem where the perennial species are long-lived, meaning compositional turnover may take many more years than we report on here.}
Our results suggest compositional shifts would have the largest impact at high rainfall because the current community maintained consistent ecosystem functional response at very low water availability.
\new{However, the increased sensitivity of ANPP to drought conditions suggests a shift in ecosystem functional response may be near, and simply take more time than we observed here.}

### 2017-12-09 manuscript/usses_water_ms.tex
more revisions; adds sensitivity analysis

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (approximately \(\pm50\%\)) and
measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the
relationship between interannual variation in available soil moisture
and interannual variation in ANPP, and if community dynamics underlie
the ecosystem responses. In particular, we are interested in the
consistency of the soil moisture-ANPP relationship among treatments. Is
the relationship steeper under the drought treatment at low soil
moisture? Does the relationship saturate under the irrigation treatment
at high soil moisture? To answer these questions we fit a generalized
linear mixed effects model to test whether the regressions differed
among treatments. We also analyzed community composition over time,
allowing us to place our experimental results within the framework our
scenarios (Fig. 1).

@@

@@

@@

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP
(Table 1; Fig. 3B). The effect of soil moisture for each treatment is
associated with high uncertainty, however, with 95\% Bayesian credible
intervals that overlap zero (Table 1). Although the parameter estimates
for the effect of soil moisture overlap zero, the posterior
distributions of the slopes all shrank and shifted to more positive
values relative to the prior distributions (Fig. A2-2), which indicates
the data did influence parameter estimates beyond the information from
the uninformative priors. Ecosystem functional response was similar
among treatments (Table 1; Fig. 3B), but there is evidence that the
slope for the irrigation treatment is less than the slope for the
control treatment. This evidence comes from interpreting the posterior
distribution of the slope offset for the irrigation treatment, from
which we calculate a \new{96}\% one-tailed probability that the estimate
is less than zero (Fig. 3A, right panel).

Community composition was similar among treatments. The multidimensional
space of community composition overlapped among treatments in all years
and was equally dispersed in all years (Table 2; Fig. 4). Community
composition was also remarkably stable over time, with no evidence of
divergence among treatments (Table 2; Fig. 4).

@@

The results of our five year experiment in a sagebrush steppe conform to
two of our four predictions, depending on treatment. Ecosystem
functional response and community composition did not change under
chronic drought (Fig. 3A, Fig. 4), representing the top left scenario in
Fig. 1. Ecosystem functional response under chronic irrigation was
different from the control treatment, but community composition remained
unchanged (Fig. 3A, Fig. 4), representing the top right scenario in Fig.
1. The decrease in the slope of the VWC \(\times\) productivity
relationship in the irrigated plots is consistent with a weaker response
of productivity to additional water in this ecosystem. This suggests
that if average soil moisture were pushed consistently higher than
currently observed ambient conditions, there would be a weaker
relationship between precipitation and productivity in this system.
Nonetheless, the slopes of the VWC \(\times\) productivity relationship
were similar among treatments (Table 1), at least at the level of
irrigation we imposed. We therefore conclude that ecosystem functional
response is consistent and weak across all precipitation treatments.

The similarity of ecosystem functional response (Fig. 3) and community
composition (Fig. 4) among treatments is surprising because grasslands
generally, and sagebrush steppe specifically, are considered
water-limited systems. \new{For example,} \citet{Huxman2004} and
\citet{Knapp2015}
\new{showed that semi-arid sites are more sensitive to drought than mesic sites, and}
\citet{Wilcox2017}
\new{found that semi-arid sites are particularly sensitive to irrigation treatments.}
\new{Based on these findings}, we expected ecosystem functional
response, community composition, or both to change under our treatments.
Why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons; two are limitations of our study, and one
is the life history traits of the species in our focal communities. We
first discuss the potential limitations of our study, and then discuss
the biological explanation.

@@

Our results suggest the species in our focal plant community are
insensitive to to changes in precipitation regime. Such insensitivity
can buffer species against precipitation variability in this semi-arid
ecosystem, making them successful in the long run. Longer, chronic
precipitation alteration might reveal plant community shifts that we did
not observe \citep[e.g.,][]{Wilcox2016}. For example, a long-term
increase in water availability could allow species that do not bet hedge
to gain prominence and dominate the ecosystem functional response.
\new{Time \emph{per se} may be especially relevant in our focal ecosystem where the perennial species are long-lived, meaning compositional turnover may take many more years than we report on here.}
Our results suggest compositional shifts would have the largest impact
at high rainfall because the current community maintained consistent
ecosystem functional response at very low water availability.

>>>>

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (approximately \(\pm50\%\)) and
measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the
relationship between interannual variation in available soil moisture
and interannual variation in ANPP, and if community dynamics underlie
the ecosystem responses. In particular, we are interested in the
consistency of the soil moisture-ANPP relationship among treatments. Is
the relationship steeper under the drought treatment at low soil
moisture? Does the relationship saturate under the irrigation treatment
at high soil moisture? To answer these questions we fit a generalized
linear mixed effects model to test whether the regressions differed
among treatments. We also analyzed community composition and
\new{the sensitivity of ANPP to drought and irrigation treatments} over
time, allowing us to place our experimental results within the framework
our scenarios (Fig. 1).

@@

\subsubsection{Ecosystem functional
response}\label{ecosystem-functional-response}

@@

\subsubsection{Sensitivity of ANPP over
time}\label{sensitivity-of-anpp-over-time}

\new{Unfortunately, our data does not allow us to include the effect of time on ecosystem functional response in the model described above. 
This is because we lack sufficient within-year and within-treatment variation of soil moisture (i.e., within a year and treatment each plot shares the same value of VWC). 
Therefore, we conducted an independent analysis to determine if the sensitivity of ANPP to treatment changed over time.}

\new{We define `sensitivity' as $\frac{ \text{ANPP}_{\text{control}} - \text{ANPP}_{\text{treatment}} } { \text{VWC}_{\text{control}} - \text{VWC}_{\text{treatment}} }$, following}
\citet{Wilcox2017}.
\new{This metric of sensitivity measures the response of ANPP standardized to the magnitude of the treatment. Because our treatment and control plots are not paired, we conducted this analysis by comparing each treatment plot to the mean of the control plots in each year. Then, independently for each treatment, we regressed sensitivity against year of treatment using the \texttt{lm()} function in R}
\citep{R2016}.
\new{This analysis also allows us to link particularly sensitive treatment years to changes in community composition.}

\subsubsection{Community composition over
time}\label{community-composition-over-time}

@@

\new{We conducted the above analysis for all species, and then conducted separate analyses for perennial species only and annual species only. Annual species have shorter life spans, so conducting separate analyses allows us to discover if community responses are restricted to short-lived species that might respond more quickly to altered precipitation regimes. Given the dominance of perennial species in our system, shifts in the annual plant community could be masked in the analysis of the full community.}

\subsubsection{Reproducibility}\label{reproducibility}

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP
(Table 1; Fig. 3B). The effect of soil moisture for each treatment is
associated with high uncertainty, however, with 95\% Bayesian credible
intervals that \new{broadly} overlap zero (Table 1). Although the
parameter estimates for the effect of soil moisture overlap zero, the
posterior distributions of the slopes all shrank and shifted to more
positive values relative to the prior distributions (Fig. A2-2), which
indicates the data did influence parameter estimates beyond the
information from the uninformative priors. Ecosystem functional response
was similar among treatments (Table 1; Fig. 3B), but there is evidence
that the \new{slopes for the drought and} irrigation treatments are less
than the slope for the control treatment. This evidence comes from
interpreting the posterior distributions of the slope offsets for the
treatments. \new{From these distributions,} we calculate a \new{91}\%
one-tailed probability that the estimate is less than zero for the
irrigation treatment
\new{and a 96\% one-tailed probability that the estimate is less than zero for the drought treatment}
(Fig. 3A, right panel).

\new{Sensitivity of ANPP to irrigation was constant over the course of our experiment (Fig. x). Sensivitity of ANPP to drought, however, grew over time (\emph{P} = 0.0002; Fig. x)}.

Community composition was similar among treatments. The multidimensional
space of community composition overlapped among treatments in all years
and was equally dispersed in all years (Table 2; Fig. 4). Community
composition was also remarkably stable over time, with no evidence of
divergence among treatments (Table 2; Fig. 4).
\new{Splitting the analysis apart for perennials and annuals led to similar results.
The perennial community was stable over time and consistently similar among treatments (Fig. Ax-x).
There is some evidence that the annual community changed in response to treatment in two years (Fig. Ax-x), but these responses appear to come from very rare species since our analysis weights them equally (Appendix X).}

@@

The results of our five year experiment in a sagebrush steppe
\new{conforms to one} of our four predictions. Ecosystem functional
response under chronic \new{drought and irrigation} was different from
the control treatment, but community composition remained unchanged
(Fig. 3A, Fig. 4), representing the top right scenario in Fig. 1. The
decrease in the slope of the VWC \(\times\) productivity relationship in
the irrigated plots is consistent with a weaker response of productivity
to additional water in this ecosystem. This suggests that if average
soil moisture were pushed consistently higher than currently observed
ambient conditions, there would be a weaker relationship between
precipitation and productivity in this system.

\new{The decrease in the slope of the VWC $\times$ productivity relationship in the droughted plots is puzzling because it is counter to our expectation that the slope would be higher than the control plots.
However, we believe the significance of the slope offsets has more to do with noise than signal.
This is because the full distributions of} the slopes of the VWC
\(\times\) productivity relationship, \new{not just the offsets}, were
similar among treatments (Table 1). We therefore conclude that ecosystem
functional response is consistent (\new{similar values}) and weak
(\new{all broadly overlapping zero}) across all precipitation
treatments.

The similarity of ecosystem functional response (Fig. 3) and community
composition (Fig. 4) among treatments is surprising because grasslands
generally, and sagebrush steppe specifically, are considered
water-limited systems. \new{For example,} \citet{Huxman2004} and
\citet{Knapp2015}
\new{showed that semi-arid sites are more sensitive to drought than mesic sites, and}
\citet{Wilcox2017}
\new{found that semi-arid sites are particularly sensitive to irrigation treatments.}
\new{Based on these findings}, we expected ecosystem functional
response, community composition, or both to change under our treatments.
Why did our treatments fail to induce ecosystem or community responses?
We can think of three reasons. Two are limitations of our study, and one
is the life history traits of the species in our focal communities. We
first discuss the potential limitations of our study, and then discuss
the biological explanation.

@@

Our results suggest the species in our focal plant community are
insensitive to to changes in precipitation regime,
\new{at least over the five years of our experiment}. Such insensitivity
can buffer species against precipitation variability in this semi-arid
ecosystem, making them successful in the long run. Longer, chronic
precipitation alteration might reveal plant community shifts that we did
not observe \citep[e.g.,][]{Wilcox2016}. For example, a long-term
increase in water availability could allow species that do not bet hedge
to gain prominence and dominate the ecosystem functional response.
\new{Time \emph{per se} may be especially relevant in our focal ecosystem where the perennial species are long-lived, meaning compositional turnover may take many more years than we report on here.}
Our results suggest compositional shifts would have the largest impact
at high rainfall because the current community maintained consistent
ecosystem functional response at very low water availability.
\new{However, the increased sensitivity of ANPP to drought conditions suggests a shift in ecosystem functional response may be near, and simply take more time than we observed here.}
